---
title: 'Lecture 8: GP demo'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(rstan)
library(mnormt)
library(shinystan)
library(plgp)
library(reshape2)
set.seed(02112021)
options(mc.cores = parallel::detectCores())
#setwd("/Volumes/GoogleDrive/My Drive/Teaching/STAT534/Lectures/Lecture 4. Stan and GP")
```



#### Conditional Normal distribution
Continuing the theme from last week, now there is one more location that we are interested in learning the temperature, maybe Rendezvous Ski Trails in West Yellowstone.

Let's assume that 
$$\begin{bmatrix} y_{bridger}\\ y_{big sky}\\ y_{rendezvous} \end{bmatrix} \sim N(\begin{bmatrix} 15 \\ 15 \\ 15\end{bmatrix}, 100\begin{bmatrix} 1 & .3 & .2 \\ .3 & 1 & .5 \\ .2 & .5 & 1
\end{bmatrix})$$

###### 1. Simulate one data point for Bridger and Big Sky

Simulate a single realization from the distribution for Bridger and Big Sky (ignoring Rendezvous for now).

###### 2. Estimate Rendezvous, conditional on the data point from Bridger and Big Sky

Conditional on the values from the previous question, construct the distribution for the Rendezvous temperature. Add this to a graph with a marginal temperature. Rerun this a few times (with a different seed) to see how it changes.

#### GP in 1D

Recall our simulated Gaussian process in 1D

```{r}
phi <- 1
sigmasq <- 1
n <- 50
x <- seq(0, 10, length.out = n)
d <- sqrt(plgp::distance(x))
eps <- sqrt(.Machine$double.eps) 
H <- exp(-d/phi) + diag(eps, n) 
y <- rmnorm(1, rep(0,n),sigmasq * H)
tibble(y = y, x = x) %>% ggplot(aes(y=y, x=x)) +
  theme_bw() + ggtitle('Random realization of a GP with phi = 1 and sigmasq = 1') +
  geom_point(size = .5)
```

We have simulated y ~ N(mu, sigmasq * H(phi)), where H(phi) is a correlation matrix from exp(-d/phi). 

##### STAN CODE 
Let's first write stan code to estimate phi, sigmasq and mu


#### GP regression in 1D

Now add a covariate

```{r}
phi <- 1
sigmasq <- 1
n <- 50
x <- seq(0, 10, length.out = n)
beta <- 1
d <- sqrt(plgp::distance(x))
eps <- sqrt(.Machine$double.eps) 
H <- exp(-d/phi) + diag(eps, n) 
y <- rmnorm(1, x * beta,sigmasq * H)
reg_fig <- tibble(y = y, x = x) %>% ggplot(aes(y=y, x=x)) +
  theme_bw() + ggtitle('Random realization of a GP with phi = 1 and sigmasq = 1') +
  geom_point(size = .5)
reg_fig
```

We have simulated y ~ N(mu, sigmasq * H(phi)), where H(phi) is a correlation matrix from exp(-d/phi). 

##### STAN CODE 
Let's first write stan code to estimate phi, sigmasq and beta


##### Making Predictions

For today, consider "plug in" estimates of phi, beta, and sigmasq. To make predictions from -1 to 11.
